\protect Connecting every bit of knowledge: \\ The structure of Wikipedia's First Link Network

### 2016-12-07 writeup/2016-04wikipedia-network/wikipedia-network-revtex4-combined.tex
accepted arXiv version

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{traversal_visit_algo_figure.pdf}  
  \caption{
    \textbf{Traversal Visit Algorithm on a sample network.}
     The traversal visit vectors are an adjacency matrix for the paths through the network: 
     the first column indicates the path formed starting with article A. 
     The number of traversal visits for article A is then the number of paths containing A 
     or the sum of the first row in our matrix:
     $\sum_{i=1}^7 A_{\text{visit, i}} = 7$.
  }
  \label{fig:Traversal Visits}
\end{figure*}

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Previous studies have used flow to characterize the structure of river networks
\cite{horton1945erosional, dodds1999unified} and describe the organization of 
blood networks, food systems, and transportation networks
\cite{garlaschelli2003universal}.

@@

While traversal visits measure accumulation, each article's first link also 
influences the shape of the FLN. 
At a point of accumulation, a single article's first link 
can exert great influence over the shape of the FLN by directing many
references on a particular path. To distinguish between an article 
that simply happened to fall within a cycle from an article funneling 
many first links, we develop a second metric called {\it traversal funnels}.
To measure traversal funnels, we traverse the FLN in the same manner as we 
did for traversal visits, but end a path once we enter a cycle.
We are then able to distinguish between an article related to many other ideas
only by virtue of its place in a cycle, from an article exerting influence over where the first links flow. 
In our simple network 
(see Fig.~\ref{fig:Traversal Funnels}), article C 
directs the flow of links towards the 3-cycle, while articles A and B are 
recipients of the flow---without exerting direct influence themselves.

By considering the FLN not only as collection of directly linked pairs of articles, but
as a flow, with accumulated references, cycles, and path-connected groups, we can explore how the many articles in Wikipedia are organized and related.

\section{Results}

\subsection{Degree Distribution}

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{articles_ndegree.pdf}
  \caption{
    \textbf{Highest Ranking Articles by in-degree.}
    We rank each article by the number of direct first links to the article (in-degree). The highest-ranking articles tend to represent geographical and biological abstractions. A full online appendix of the results and data is available \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.}
  \label{fig:indegree list}
\end{figure}

We rank all 11 million articles in the FLN by in-degree to find 
the ``United States'' with 80,249 direct first links as the most referenced
Wikipedia article 
(Fig.~\ref{fig:indegree list}). 
Other high-ranking articles
include foundational abstract concepts such as ``village,'' ``species,''; 
sports associations such as ``American Football,'' ``Association Football''; 
and developed nations such as ``France,'' ``Japan,'' ``Russia,'' ``Australia,'' and 
the ``Netherlands.'' These high ranking articles are useful abstractions: nations
describe a collection of individuals with a common culture, language, or 
geographical proximity; sports teams describe an ever changing collection of 
players often associated with a cultural identity or a geographical 
region. 
Since abstractions such as nations and teams are inherently comprised
of many parts, authors often reference the abstraction when describing a part.

>>>>

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{traversal_visit_algo_figure.pdf}  
  \caption{
    \textbf{Traversal Visit Algorithm on a sample network.}
     The traversal visit vectors are an adjacency matrix for the paths through the network: 
     the first column indicates the path formed starting with article A. 
     The number of traversal visits for article A is then the number of paths containing A 
     or the sum of the first row in our matrix:
     $\sum_{i=1}^7 A_{\text{visit, i}} = 7$.
  }
  \label{fig:Traversal Visits}
\end{figure*}
Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Previous studies have used flow to characterize the structure of river networks
\cite{horton1945erosional, dodds1999unified} and describe the organization of 
blood networks, food systems, and transportation networks
\cite{garlaschelli2003universal}. 
In the same vein, we use flow to measure accumulation and develop a new method for isolating influence in a directed network with cycles.

\subsection{Traversal Visits}
\label{Traversal Visits}

@@

\end{figure}
\subsection{Traversal Funnels}
\label{Traversal Funnels}

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{centrality_measures.pdf}
  \caption{
    \textbf{Centrality Measures}
    The table above compares standard centrality measures on the sample network against traversal funnels. While betweenness centrality and eigenvector centrality both reward node A for its place in the cycle, traversal funnels only rewards node C for directing paths into the cycle. }
  \label{fig:Centrality Measures}

While traversal visits measure accumulation, each article's first link also 
influences the shape of the FLN. 
At a point of accumulation, a single article's first link 
can exert great influence over the shape of the FLN by directing many
references on a particular path. To distinguish between an article 
that simply happened to fall within a cycle from an article funneling 
many first links, we develop a second metric called {\it traversal funnels}.
To measure traversal funnels, we traverse the FLN in the same manner as we 
did for traversal visits, but end a path once we enter a cycle.
We are then able to distinguish between an article related to many other ideas
only by virtue of its place in a cycle, from an article exerting influence over where the first links flow. 
In the sample network 
(see Fig.~\ref{fig:Traversal Funnels}), article C 
directs the flow of links towards the 3-cycle, while articles A and B are 
recipients of the flow---without exerting direct influence themselves.

The algorithm increments the traversal funnels count along each path, up to the start of a cycle. The resulting count is each node's number of traversal funnels: the number of paths the node directs up to the start of a cycle. Below is the pseudo-code.

\begin{verbatim}
Traversal Funnels Algorithm

// 1. Determine Cycles 
for node in Graph
    next_node = node.link
    visited = {node, next_node}
    // advance to potential cycle
    while next_node not in visited:
        next_node = next_node.link
        visited.add(next_node)
    // mark cycle
    if (node == next_node):
        node.in_cycle = True

// 2. Compute Traversal Funnels
for node in Graph:
    while node.in_cycle == False:
        node.funnels += 1
        node = node.link
\end{verbatim}

The algorithm's average runtime is proportional to the product of the number of nodes, $n$, and the average path length, $l$. If the average path length is fixed (independent of $n$), the runtime is $O(n)$ with $O(n + l)$ auxiliary memory for storing cycles and traversing paths. In the worst case where $l$ is proportional to $n$, the runtime is $O(n^2)$ with $O(n)$ auxiliary memory for storing cycles and traversing paths. In implementation, each iteration step in the algorithm can be parallelized since the algorithm is order agnostic, as long as each node is eventually selected.

As a measure of influence, traversal funnels differs from traditional centrality measures such as degree centrality, betweenness centrality, and eigenvector centrality. 
Degree centrality characterizes how many articles have a first link to a particular topic, restricting influence to a single link. 
Betweenness centrality and eigenvalue centrality capture flow within the network more than a single link away. 
Unlike traversal funnels however, betweenness centrality and eigenvalue centrality reward nodes that simply happen to be in a cycle. 
We compute betweenness centrality and eigenvalue centrality on the sample network using \href{https://networkx.github.io/}{NetworkX}, a Python package for modelling complex networks (see the \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{online appendix} for details). 
In our sample network (see Fig.~\ref{fig:Centrality Measures}), node A has betweenness centrality of 0.17 and eigenvalue centrality of 0.58, although node A is simply the recipient of paths directed by node C.
Eigenvector centrality even assigns the same centrality to nodes A, B, and C.

In contrast, traversal funnels defines a path without a predetermined end node, following the progression of links to its culmination at the start of a cycle. Furthermore, traversal funnels provides an interpretable metric: a node's traversal funnels is the number of paths the node directs to the start of a cycle. 
Consequently, traversal funnels allow us to isolate the paths directed by node C, of which there are 4, without any spill over effect on other nodes in its cycle: A and B have zero traversal funnels.

In addition, traversal funnels extends to directed networks with more than one outward edge. In such a network, there may be multiple paths associated with a single starting node. 
The same interpretation for traversal funnels as the number of paths a node directs to a cycle still holds. 
In contrast to betweenness centrality, traversal funnels would not favor the shortest path over any other. Traversal funnels equally weighs all paths up to a cycle.

By considering the FLN not only as collection of directly linked pairs of articles, but
as a flow using traversal funnels, with accumulated references, cycles, and path-connected groups, we explore how the many articles in Wikipedia are organized and related.

\section{Results}
\subsection{Degree Distribution}

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{articles_ndegree.pdf}
  \caption{
    \textbf{Highest Ranking Articles by in-degree.}
    We rank each article by the number of direct first links to the article (in-degree). The highest-ranking articles tend to represent geographical and biological abstractions. A full online appendix of the results and data is available \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.}
  \label{fig:indegree list}
\end{figure}
We rank all 11 million articles in the FLN by in-degree to find 
the ``United States'' with 80,249 direct first links as the most referenced
Wikipedia article 
(Fig.~\ref{fig:indegree list}). 
Other high-ranking articles
include foundational abstract concepts such as ``village,'' ``species,''; 
sports associations such as ``American Football,'' ``Association Football''; 
and developed nations such as ``France,'' ``Japan,'' ``Russia,'' ``Australia,'' and 
the ``Netherlands.'' These high ranking articles are useful abstractions: nations
describe a collection of individuals with a common culture, language, or 
geographical proximity; sports teams describe an ever changing collection of 
players often associated with a cultural identity or a geographical 
region. 
Since abstractions such as nations and teams are inherently comprised
of many parts, authors often reference the abstraction when describing a part.
